Historical aerial surveys map long-term changes of forest cover and structure in the central Congo Basin

### 2020-01-24 manuscript/main.tex
revision

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{Land Use and Land Cover change (LULCC) of African rainforest contribute
to global carbon emissions. Yet, most historical estimates of LULCC and
their carbon emissions rely on non-spatially explicity data in the
pre-satellite era (\textless{} 1972). Past studies have expanded this
pre-satellite time frame with declassified satellite surveillance data
or aerial surveys, but none cover the Congo Basin. Here, we use
historical aerial survey photos to map the extent and structure of LULCC
around Yangambi, central Congo Basin, in 1958. Our study leveraged
Structure-from-Motion to generate a large orthomosaic covering 82800 ha,
counting 733 million pixels, at a sub meter resolution and
geo-referenced to \textasciitilde{}4.7 \(\pm\) 4.3 m. Primary forest in
the orthomosaic was classified with a Convolutional Neural Network Deep
Learning approach. We used these data to quantify LULCC, landscape and
canopy complexity and scale above ground biomass between historical and
contemporary condition. We show a shift from historical highly
structured industrial deforestation in 1958 (16200 ha), to contemporary
smallholder farming clearing (8800 ha), increasing landscape
fragmentation and forest edges. Efforts to quantify canopy texture and
their link to carbon storage had limited to no success. Our analysis
provides an insight in key spatial and temporal patterns of
deforestation and reforestation at a multi-decadal scale and provide a
historical context to land-cover and land-use change spatial patterns
for past and ongoing field research in the area.}

@@

Tropical ecosystem services are severely impacted by deforestation and
forest degradation \citep{hansen2013, houghton2012, tyukavina2015}. Not
only does tropical forest Land-Use and Land-Cover Change (LULCC)
constitute 10 to 15\% of the total global emissions
\citep{vanderwerf2009}, changes in forest fragmentation affect the
forest structure and function. Strong fragmentation effects decreases
the number of large trees along forest edges
\citep{laurance2000, magnago2015}, while species composition and
biodiversity are equally negatively affected
\citep{poorter2006, barlow2016, vandeperre2018}. Estimates show that
31\% of carbon emissions are caused by edge effects alone
\citep{brinck2017}.

As such, anthropogenic factors through LULCC heavily affect forest
structure and functioning \citep{fauset2017, brinck2017, didham1999}.
Accurate estimates of LULCC and forest canopy structure are therefore
imperative to estimate carbon emissions and other ecosystem services
\citep{houghton2012, hansen2013}. Remote sensing products have been key
input as they provide accurate spatial information to help estimate
carbon emissions \citep{mitchard2018, hansen2013}. High resolution
aerial images provide scientists tools to monitor forest extent,
structure and carbon emissions as canopy texture is linked to above
ground biomass \citep{Ploton2012, Couteron2005, Barbier2010}. Yet, these
estimates are often limited in time to recent decades
\citep{defries2002, houghton2012, hansen2013, achard2014}.

Historical estimates \textbf{of Land-Use and Land-Cover (LULC)}, in the
pre-satellite era (\textless{} 1972), exist but generally rely on
non-spatially explicit data (i.e socio-economic data)
\citep{houghton2012, ramankutty1999, defries2002, sader1988}. Efforts
have been made to use other geo-spatial data sources such as historical
maps \citep{willcock2016}, declassified CORONA satellite surveillance
data across the US and central Brazil \citep{song2015} as well as aerial
surveys in post world war II German \citep{nita2018}. Survey data across
the African continent is less common, inaccessible or both. Some studies
do exist, as \citet{buitenwerf2012}, \citet{hudak1998} and
\citet{okeke2006} used aerial survey images to map vegetation changes in
South African savannas and the Adulam Nature Reserve, respectively,
whilst \citet{frankl2015} and \citet{nyssen2016} mapped the Ethiopian
highlands of 1930's.

Despite the existence of large historical archives of aerial survey
imagery for the central Congo Basin (Figure 1, Appendix Figure 3) as of
yet no studies have valorized these data. The lack of a consistent
valorization effort is unfortunate as the African rainforest is the
second largest on Earth and covers \textasciitilde{}630 million ha,
representing up to 66 Pg of carbon storage \citep{Lewis2009b}, and
currently loses forest at an increasing pace \citep{butsic2015}. Given
the impact of LULCC on the structure and functioning of central African
tropical forests, and their influence on both carbon dynamics
\citep{Kearsley2013} and biodiversity \citep{vandeperre2018}, accurate
long term reporting of historical forest cover warrants more attention
\citep{willcock2016}.

Here, we use a combination of historical aerial photography and
contemporary remote sensing data to map long-term changes in the extent
and structure of the tropical forest surrounding Yangambi, in the
central Congo Basin, effectively linking the start of the antrophocene
\citep{lewis2015} with current assessments. Yangambi was, and remains, a
focal center of forest and agricultural research and development in the
central Congo Basin. Past research in the region allows for thorough
assessment of land-use and land-cover change from a multi-disciplinary
point of view, confronting us with complex deforestation and land-use
patterns.

@@

Data for the central Congo Basin region, surrounding Kisangani, were
collected in several flights during the dry season of 1958 and 1959
(from 8/01/1958 to 20/02/1958 and from 28/12/1958 to 9/01/1959
respectively, see Appendix Figure 1) to generate topographic maps of the
area, supervised by the ``Institut Géographique du Congo Belge'' in
Kinshasa (then Léopoldville). Black-and-White infrared images (0.4 - 0.9
\(\mu\)m) were gathered along flight paths running mostly from west to
east, between \textbf{09:00 and 11:00} local time. Along a flight path
continuous images were taken using a Wild Heerburgg RC5a (currently
Leica Geosystems) with an Aviogon lens assembly (114.83mm / f 5.6, with
a 90\degree view angle) resulting in square photo negative of 180 by 180
mm. Flights were flown at an average absolute altitude of
\textasciitilde{}5200 m above sea level, covering roughly 18 530
km\textsuperscript{2} at an approximate scale of 1/40 000. The use of
the integrated autograph system ensured timely acquisition of pictures
with a precise overlap (\textasciitilde{}1/3) between images. This large
overlap between images together with flight parameters would allow
post-processing, using stereographs, to create accurate topographic
maps. Original data from this campaign are stored in the
\href{https://www.africamuseum.be/en}{Royal Museum for Central Africa}
in Tervuren, Belgium (Figure 1).

@@

\caption{A box of historical aerial photographs (left) and a single aerial photograph (right) showing part of the Congo river. Note the meta-data provided in the margins of the image such as acquisition time and fleight height and the solar glare on the Congo river.}\label{fig:unnamed-chunk-1}
\end{figure}

@@

Data was processed into a georeferenced orthomosaic using a Structure
from motion (SfM, \citet{ullman1979}) approach implemented in
\href{http://www.agisoft.com}{Agisoft Metashape} version 1.5.2 (Agisoft
LLC, St.~Petersburg, Russia). An orthomosaic corrects remote sensing
data to represent a perfectly downward looking image, free from
perspective distortions due to topography and camera tilt. Using the SfM
technique features, areas in images with a large degree of similarity,
are matched across various images to reconstruct a three dimensional
scene (topography) from two-dimensional image sequences. During the SfM
analysis we masked clouds, glare or large water bodies such as the Congo
river.

@@

The CNN model was trained for 100 epochs with a batch size of 30 using
Adam optimization \citep{kingma2017}, maximizing the
Intersect-over-Union (IoU) using Sørensen--Dice and categorical
cross-entropy loss functions. Data augmentation included random cropping
to 320 pixel squares, random orientation, scaling, perspective, contrast
and brightness shifts and image blurring. The optimized model was used
to classify the complete orthomosaic using a moving window approach with
a step size of 110 pixels and a majority vote \textbf{(\textgreater{}
50\% agreement)} across overlapping areas to limit segmentation edge
effects. \textbf{In addition we provide, raw pixel level classification
agreement data for quality control purposes (see Data availability
below).} We refer to Figure 4 for a synoptic overview of the full deep
learning learning workflow.

@@

To map long term land-use and land-cover change in the Yangambi region
we used the contemporary Global Forest Change version 1.6 (GFC, tile
10N-020E) map data \citep{hansen2013}. Using the \textbf{GFC} data we
calculated the latest state of the forest with respect to the conditions
at the start of 1958, 60 years earlier. In our analysis we only included
\textbf{GFC} pixels which recorded no \textbf{forest} loss throughout
the whole 2000 - 2018 period. As the resolution of the historical forest
classification exceeds that of the GFC map we downsampled our historical
forest cover data to 30 m GFC resolution \textbf{using a nearest
neighbour approach}. We masked out all water bodies using the Global
Forest Change survey data mask layer, and limited the analysis to the
right bank of the Congo river. We provide summary statistics of
historical and contemporary forest loss and regrowth.

deforestation (persistent forest loss after 1958), reforestation
(regrowth after loss in 1958).

@@

We estimated above ground carbon (AGC) losses and gains over time using
plot based averages of recent inventory data at permanent sampling plots
in the area (Figure 2). We refer to \citet{Kearsley2013} for the survey
method and allometric relations used to scale the survey data. Unlike
standard square 1 ha plots edge plots were set back 200m from forest
edges and were 50x200 m, with the 50 m side of the plot along the forest
edge and continuing 200 m into the forest (Appendix Table 1). We further
confirmed that forest edge plots did not show a significantly different
AGC compared to those of non-edged / mixed forest plots (Mann Whitney U
test, p \textless{} 0.05). Thus it was not necessary to explicitly
quantify changes in AGC caused by edge effects. Moreover, we used the
mean values of the mixed forest as representative for potential AGC
losses. Despite the challenges inherent in quantifying AGC for forest
edges we mapped the total extent of the edges in the contemporary
landscape. To align our landscape analysis with exploratory analysis of
the survey data we used a buffer of 200 m to estimate the extent of
forest edges and patches, up to the location of forest edge plots.

Surveys of old plantations show a large variation in AGC, depending on
age and the crop type. For example, the AGC values varied from 168.67 to
86.55 g Mg C ha\textsuperscript{-1}, for \emph{Hevea brasiliensis}
(rubber tree) and \emph{Elaeis guineensis} (oil palm) plots respectively
(\citet{bustillo2018}, personal communications). These higher values are
in line with the mixed AGC estimates in the area, while the palm
plantations resemble old-regrowth values (81.8 Mg C
ha\textsuperscript{-1}, see \textbf{Appendix Table 1}). We therefore use
both the estimates of old-regrowth and mixed forest to estimate AGC for
regrowth. We did not have sufficient data to account for individual
changes in AGC across plantations.

@@

We compared the structure of the canopy both visually and using Fourier
Transform Textural Ordination (FOTO, \citet{couteron2002}). Fourier
Transform Textural Ordination uses a principal component analysis (PCA)
on radially averaged 2D Fourier spectra to characterize canopy (image)
texture. The FOTO technique was first described by \citet{couteron2002}
to quantify canopy stucture in relation to biomass and biodiversity, and
can be used across multiple scenes using normalization
\citep{Barbier2010}.

@@

For both site based and scene analysis we correlate PC values with
permanent sample plot inventory data such as stem density, above ground
biomass and tree species richness. Additional comparisons are made
between contemporary Geo-Eye data and the historical orthomosaic derived
PC values. Due to the few available permanent sampling plots in both
scenes we used a non-parametric paired signed rank (Wilcoxon) test to
determine differences between the PC values of the Geo-Eye and
historical orthomosaic image scenes across mono-dominant and mixed
forest types. In all analysis, mono-dominant site 4 was removed from the
analysis due to cloud contamination.

@@

Our analysis provides a first spatially explicity historical composite
of aerial survey images \textbf{in support of mapping land-use and
land-cover} within the Congo Basin. The use of high resolution
historical images combined with structure from motion image processing
techniques allowed us to mosaic old imagery across a large extent. The
final orthomosaic composition of the Yangambi region \textbf{provided}
an image scene covering approximately 733 million pixels across
\textbf{\textasciitilde{}82800 ha} (\textasciitilde{}23x36 km, Figure
2). The overall accuracy of the SfM orthomosaic composition was 0.88
m/pixel using the sparse cloud DEM for corrections at 45.8 m/pixel. The
resulting georeferenced scene had a spatial accuracy of approximately
\textbf{23 m}. Further georeferencing outside the SfM workflow reduced
the mean error at the ground control points to 5.3 \(\pm\) 4.9 px
(\textasciitilde{}4.7 \(\pm\) 4.3 m), with a median error of 2.9 px (2.6
m). The orthomosaic served as input for all subsequent LULCC analysis
with all derived maps provided with the manuscript repository (see data
\& code availability statements below).

@@

\textbf{The CNN deep learning classifier reached an
Intersection-over-Union (IoU) accuracy of 97\% on the detection of
disturbed forest in the out-of-sample (validation) synthetic landscape
data. Using all pixels within the validatoin polygons (Figure 5) showed
a similar accuracy value of \textasciitilde{}98\%. Using all polygons
across the scene, including those used in the generation of testing and
training synthetic landscapes, increased the accuracy to
\textasciitilde{}99\% (Table 2). A comparison with recent pan-chromatic
Geo-Eye data shows good agreement, with an accuracy of
\textasciitilde{}87\% across all pixels, between the landsat based GFC
data and downscaled CNN results (Table 2 and Figure 7).}

@@

\hypertarget{long-term-lulc-changes-and-above-ground-carbon}{%
\subsection{Long term LULC changes and Above Ground
Carbon}\label{long-term-lulc-changes-and-above-ground-carbon}}

@@

Recent deforestation follows a distincly different pattern compared to
historical patterns. Historical deforestation showed a classical
fishbone pattern for forest clearing with very sharp edges, while
current patterns are patchy and ad-hoc (\textbf{figure 6C}, dark blue
and green colours respectively). These differences are reflected in the
analysis of landscape metrics of forest loss. Between the historical and
contemporary LULCC maps we see an increase in small disturbances, as
indicated by the decreasing area of the mean patch size, down to
\textasciitilde{}1.86 \(\pm\) 0.75 ha from \textasciitilde{}5.25 \(\pm\)
5.02 ha historically. Perimeter lengths were longer historically, at
1451 \(\pm\) 943 m, compared to contemporary landscapes
\textasciitilde{}921 \(\pm\) 362 m (Table 3). This shift in perimeter
area ratio led to a similar change in the fractal index, increasing in
value to 1.1 \(\pm\) 0.05 from 1.09 \(\pm\) 0.04 over time. Values
closer to a fractal index of 2 suggest a more complex (fragmented)
landscape.

@@

Changes in both land-use and land-cover led to concomitant changes in
above ground carbon stocks. Recovery throughout the region was
characterized for patches of forest and plantations. Assuming high
density stands, based on previous work, this could amount to a carbon
gains of 1592 Gg C across our study area, offsetting more recent losses
of approximately 1408 Gg C. On the other hand, at the low end, if we
assume a lower carbon density of 81.8 Mg C ha\textsuperscript{-1} this
would result in a total carbon gain of 811 Gg C. Using our approach
results indicate that overall deforestation around Yangambi has resulted
in a loss of \textasciitilde{}2416 Gg C in AGC stocks.

@@

\caption{\label{tab:unnamed-chunk-10}Land use land cover change statistics of forest cover around Yangambi in the central Congo Basin. The data evaluates a difference between a historical (1958) aerial photography based survey and the Hansen et al. 2013 based satellite remote sensing data. Spatial coverage statistics are provided in square kilometers (km) and hectares (ha), rounded to the nearest integer as well as Above Ground Carbon (AGC) scaled using recent survey measurements.}
\centering
\begin{tabular}[t]{lrl}
\toprule
\multicolumn{2}{c}{ } & \multicolumn{1}{c}{AGC} \\
\cmidrule(l{3pt}r{3pt}){3-3}
  & ha & (Gg C $ha\textsuperscript{-1}$\\
\midrule
Forest & 68455 & \\
-  of which edges & 13151 & \\
Regrowth > 1958 loss & 9918 & 811 - 1592\\
Loss > 2000 & 8776 & 1408\\
Loss > 1958 (permanent) & 6282 & 1008\\
\bottomrule
\end{tabular}
\end{table}

@@

Visual interpretation of the scenes provide evidence that most locations
do not change dramatically with respect to canopy composition, except
for the large areas of disturbances in contemporary fallow or
young-regrowth plots. One marked difference is noted in the
mono-dominant site 6 (Appendix Table 1). Here, the current mono-dominant
\emph{Brachystegia laurenttii} is a recent development, changing the
canopy structure visibly during the last half century (\textbf{Figure
8}). The previous varied canopy structure gave way to a more dense and
uniform canopy. This is reflected in a change of the FOTO PC value from
0.19 historically to its current value of 0.54. This historical value is
similar to the mean of contemporary mono-dominant stands with PC
averaging 0.34 \(\pm\) 0.1, and is only slightly higher than historical
values for a mixed forest (0.18 \(\pm\) 0.08). The reverse pattern is
seen in the contemporary PC values. Here, the value of 0.54 exceeds
those of most mono-dominant stands (0.35 \(\pm\) 0.08), and is even
further removed from the values noted for mixed forests (0.12 \(\pm\)
0.03).

@@

Using only small subsets around existing permanent sampling plots we
show distinct differences between forest types, with PC values in both
historical and contemporary imagery markedly higher for the
mono-dominant forest types compared to all others (\textbf{Appendix
Figure 4}). Provided that the young-regrowth and fallow permanent
sampling plots have seen recent disturbance the Wilcoxon signed rank
test on the mixed and mono-dominant plots between the historical and
contemporary PC values did not show a significant difference (p
\textgreater{} 0.05). Similarly, no significant different using PC
values extracted from the whole scene analysis was noted (p
\textgreater{} 0.05). Any relationships between contemporary Geo-Eye
data and permanent sampling plot measurements of Above Ground Biomass,
stem density and species richness were non-significant (p \textgreater{}
0.05, Appendix \textbf{Figures 5-7}).

@@

\hypertarget{land-use-land-cover-change-classification-validation}{%
\subsubsection{Land-use, Land-Cover change classification \&
validation}\label{land-use-land-cover-change-classification-validation}}

@@

\hypertarget{land-use-land-cover-change-above-ground-carbon}{%
\subsection{Land-use, Land-Cover change \& Above Ground
Carbon}\label{land-use-land-cover-change-above-ground-carbon}}

>>>>

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{Given the impact of tropical forest disturbances on atmospheric carbon
emissions, biodiversity and ecosystem productivity accurate long term
reporting of Land-Use and Land-Cover (LULC) change in the pre-satellite
era (\textless{} 1972) is an imperative. Here, we use a combination of
historical (1958) aerial photography and contemporary remote sensing
data to map long-term changes in the extent and structure of the
tropical forest surrounding Yangambi (DR Congo), in the central Congo
Basin. Our study leveraged Structure-from-Motion and a novel
Convolutional Neural Network based LULC classifier, using synthetic
landscapes based image augmentation, to map historical forest cover
across a large sub meter resolution orthomosaic (\textasciitilde{}82800
ha) geo-referenced to \textasciitilde{}4.7 \(\pm\) 4.3 m. A comparison
with contemporary LULC data showed a shift from previously highly
regular industrial deforestation of large areas, to discrete smallholder
farming clearing, increasing landscape fragmentation but also providing
opportunties for substantial forest regrowth. Efforts to quantify long
term canopy texture changes and a link to above ground carbon had
limited to no success. Our analysis provides methods and insights into
key spatial and temporal patterns of deforestation and reforestation at
a multi-decadal scale, providing a historical context for past and
ongoing forest research in the area.}

@@

Tropical ecosystem services are severely impacted by deforestation and
forest degradation \citep{hansen2013, houghton2012, tyukavina2015}. Not
only does tropical forest Land-Use and Land-Cover Change (LULCC)
constitute 10 to 15\% of the total global carbon emissions
\citep{vanderwerf2009}, changes in forest fragmentation affect the
forest structure and function
\citep{fauset2017, brinck2017, didham1999}. Strong fragmentation effects
decreases the number of large trees along forest edges
\citep{laurance2000, magnago2015}, while species composition and
biodiversity are equally negatively affected
\citep{poorter2006, barlow2016, vandeperre2018}. Estimates show that
31\% of carbon emissions are caused by edge effects alone
\citep{brinck2017}.

Accurate estimates of LULCC and forest canopy structure are therefore
imperative to estimate carbon emissions and other ecosystem services
\citep{houghton2012, hansen2013}. Remote sensing products have been key
inputs in LULCC assessments as they provide accurate spatial information
to help estimate carbon emissions \citep{mitchard2018, hansen2013}. More
so, high resolution aerial images provide scientists tools to monitor
forest extent, structure and carbon emissions as canopy texture is
linked to above ground biomass
\citep{Ploton2012, Couteron2005, Barbier2010}. Yet, most of these
estimates are limited in time to recent decades
\citep{defries2002, houghton2012, hansen2013, achard2014}.

Historical estimates \textbf{of Land-Use and Land-Cover (LULC)}, in the
pre-satellite era (\textless{} 1972), exist but generally rely on
non-spatially explicit data (i.e socio-economic data)
\citep{houghton2012, ramankutty1999, defries2002, sader1988}. Efforts
have been made to use other geo-spatial data sources such as historical
maps \citep{willcock2016}, declassified CORONA satellite surveillance
data across the US and central Brazil \citep{song2015} as well as aerial
surveys in post World War II German \citep{nita2018}. Survey data across
the African continent is less common, inaccessible or both. Some studies
do exist, as \citet{buitenwerf2012} and \citet{hudak1998} used aerial
survey images to map vegetation changes in South African savannas,
whilst \citet{frankl2015} and \citet{nyssen2016} mapped the Ethiopian
highlands of 1930's.

Across the central Congo Basin most of these historical images were
collected within the context of national cartographic efforts by the
``Institut Géographique du Congo Belge'' in Kinshasa (then
Léopoldville), DR Congo. Despite the existence of large archives of
aerial survey imagery (Figure 1, Appendix Figure 3), as of yet, no
studies have valorized these data. The lack of a consistent valorization
effort is unfortunate as the African rainforest is the second largest on
Earth and covers \textasciitilde{}630 million ha, representing up to 66
Pg of carbon storage \citep{Lewis2009b}, and currently loses forest at
an increasing pace \citep{butsic2015}. Given the impact of LULCC on the
structure and functioning of central African tropical forests, and their
influence on both carbon dynamics \citep{Kearsley2013} and biodiversity
\citep{vandeperre2018}, accurate long term reporting of historical
forest cover warrants more attention \citep{willcock2016}.

Here, we use a combination of historical aerial photography and
contemporary remote sensing data to map long-term changes in the extent
and structure of the tropical forest surrounding Yangambi (DR Congo), in
the central Congo Basin, effectively linking the start of the
antrophocene \citep{lewis2015} with current assessments. Yangambi was,
and remains, a focal center of forest and agricultural research and
development in the central Congo Basin. Past research in the region
allows for thorough assessment of land-use and land-cover change from a
multi-disciplinary point of view, confronting us with complex
deforestation and land-use patterns.

@@

Data for the central Congo Basin region, surrounding Kisangani, were
collected in several flights during the dry season of 1958 and 1959
(from 8/01/1958 to 20/02/1958 and from 28/12/1958 to 9/01/1959
respectively, see Appendix Figure 1) to generate topographic maps of the
area, supervised by the ``Institut Géographique du Congo Belge'' in
Kinshasa, DR Congo (then Léopoldville). Black-and-White infrared images
(0.4 - 0.9 \(\mu\)m) were gathered along flight paths running mostly
from west to east, between \textbf{09:00 and 11:00} local time. Along a
flight path continuous images were taken using a Wild Heerburgg RC5a
(currently Leica Geosystems) with an Aviogon lens assembly (114.83mm / f
5.6, with a 90\degree view angle) resulting in square photo negatives of
180 by 180 mm. Flights were flown at an average absolute altitude of
\textasciitilde{}5200 m above sea level, covering roughly 18 530
km\textsuperscript{2} at an approximate scale of 1/40 000. The use of
the integrated autograph system ensured timely acquisition of pictures
with a precise overlap (\textasciitilde{}1/3) between images. This large
overlap between images together with flight parameters would allow
post-processing, using stereographs, to create accurate topographic
maps. Original data from this campaign are stored in the
\href{https://www.africamuseum.be/en}{Royal Museum for Central Africa}
in Tervuren, Belgium (Figure 1).

@@

\caption{A box of historical aerial photographs (left) and a single aerial photograph (right) showing part of the Congo river. Note the meta-data provided in the right side margin of the image, such as acquisition time and fleight height.}\label{fig:unnamed-chunk-1}
\end{figure}

@@

Data was processed into a georeferenced orthomosaic using a
Structure-from-motion (SfM, \citet{ullman1979}) approach implemented in
\href{http://www.agisoft.com}{Agisoft Metashape} version 1.5.2 (Agisoft
LLC, St.~Petersburg, Russia). An orthomosaic corrects remote sensing
data to represent a perfectly downward looking image, free from
perspective distortions due to topography and camera tilt. Using the SfM
technique features, areas in images with a large degree of similarity,
are matched across various images to reconstruct a three dimensional
scene (topography) from two-dimensional image sequences. During the SfM
analysis we masked clouds, glare or large water bodies such as the Congo
river.

@@

The CNN model was trained for 100 epochs with a batch size of 30 using
Adam optimization \citep{kingma2017}, maximizing the
Intersect-over-Union (IoU) using Sørensen--Dice \citep{dice1945} and
categorical cross-entropy loss functions. Data augmentation included
random cropping to 320 pixel squares, random orientation, scaling,
perspective, contrast and brightness shifts and image blurring. The
optimized model was used to classify the complete orthomosaic using a
moving window approach with a step size of 110 pixels and a majority
vote \textbf{(\textgreater{} 50\% agreement)} across overlapping areas
to limit segmentation edge effects. \textbf{In addition, we provide raw
pixel level classification agreement data for quality control purposes
(see Data availability below).} We refer to Figure 4 for a synoptic
overview of the full deep learning workflow.

@@

To map long term LULCC in the Yangambi region we used the contemporary
Global Forest Change version 1.6 (GFC, tile 10N-020E) map data
\citep{hansen2013}. Using the \textbf{GFC} data we calculated the latest
state of the forest with respect to the conditions at the start of 1958,
60 years earlier. In our analysis we only included \textbf{GFC} pixels
which recorded no \textbf{forest} loss throughout the whole 2000 - 2018
period. As the resolution of the historical forest classification
exceeds that of the GFC map we downsampled our historical forest cover
data to 30 m GFC resolution \textbf{using a nearest neighbour approach}.
We masked out all water bodies using the Global Forest Change survey
data mask layer, and limited the analysis to the right bank of the Congo
river. We provide summary statistics of historical and contemporary
forest loss and regrowth. \textbf{We map persistent forest loss after
1958, regrowth after loss in 1958, recent forest loss and long term
(stable) forest cover.}

@@

We estimated above ground carbon (AGC) losses and gains over time using
plot based averages of recent inventory data at permanent sampling plots
in the area (Figure 2). We refer to \citet{Kearsley2013} for the survey
method and allometric relations used to scale the survey data. Unlike
standard square 1 ha plots, edge plots were set back 200 m from forest
edges and were 50x200 m, with the 50 m side of the plot along the forest
edge and continuing 200 m into the forest (Appendix Table 1). We further
confirmed that forest edge plots did not show a significantly different
AGC compared to those of non-edged / mixed forest plots (Mann Whitney U
test, p \textless{} 0.05). Thus it was not necessary to explicitly
quantify changes in AGC caused by edge effects. Moreover, we used the
mean values of the mixed forest as representative for potential AGC
losses. Despite the challenges inherent in quantifying AGC for forest
edges we mapped the total extent of the edges in the contemporary
landscape. To align our landscape analysis with exploratory analysis of
the survey data we used a buffer of 200 m to estimate the extent of
forest edges and patches, up to the location of forest edge plots.

Surveys of old plantations show a large variation in AGC, depending on
age and the crop type. For example, the AGC values varied from 168.67 to
86.55 Mg C ha\textsuperscript{-1}, for \emph{Hevea brasiliensis} (rubber
tree) and \emph{Elaeis guineensis} (oil palm) plots respectively
(\citet{bustillo2018}, personal communications). These higher values are
in line with the mixed AGC estimates in the area, while the palm
plantations resemble old-regrowth values (81.8 Mg C
ha\textsuperscript{-1}, see \textbf{Appendix Table 1}). We therefore use
both the estimates of old-regrowth and mixed forest to estimate AGC for
regrowth. We did not have sufficient data to account for individual
changes in AGC across plantations.

@@

We compared the structure of the canopy both visually and using Fourier
Transform Textural Ordination (FOTO, \citet{couteron2002}). FOTO uses a
principal component analysis (PCA) on radially averaged 2D Fourier
spectra to characterize canopy (image) texture. The FOTO technique was
first described by \citet{couteron2002} to quantify canopy stucture in
relation to biomass and biodiversity, and can be used across multiple
scenes using normalization \citep{Barbier2010}.

@@

For both site based and scene analysis we correlated PC values with
permanent sample plot inventory data such as stem density, above ground
biomass and tree species richness. Additional comparisons are made
between contemporary Geo-Eye data and the historical orthomosaic derived
PC values. Due to the few available permanent sampling plots in both
scenes we used a non-parametric paired signed rank (Wilcoxon) test
\citep{wilcoxon1945} to determine differences between the PC values of
the Geo-Eye and historical orthomosaic image scenes across mono-dominant
and mixed forest types. In all analysis, mono-dominant site 4 was
removed from the analysis due to cloud contamination.

@@

Our analysis provides a first spatially explicity historical composite
of aerial survey images \textbf{in support of mapping land-use and
land-cover} within the Congo Basin. The use of high resolution
historical images combined with SfM image processing techniques allowed
us to mosaic old imagery across a large extent. The final orthomosaic
composition of the Yangambi region \textbf{provided} an image scene
covering approximately 733 million pixels across
\textbf{\textasciitilde{}82800 ha} (\textasciitilde{}23x36 km, Figure
2). The overall accuracy of the SfM orthomosaic composition was 0.88
m/pixel using the sparse cloud DEM for corrections at 45.8 m/pixel. The
resulting georeferenced scene had a spatial accuracy of approximately
\textbf{23 m}. Further georeferencing outside the SfM workflow reduced
the mean error at the ground control points to 5.3 \(\pm\) 4.9 px
(\textasciitilde{}4.7 \(\pm\) 4.3 m), with a median error of 2.9 px (2.6
m). The orthomosaic served as input for all subsequent LULCC analysis
with all derived maps provided with the manuscript repository (see data
\& code availability statements below).

@@

\textbf{The CNN deep learning classifier reached an
Intersection-over-Union accuracy of 97\% on the detection of disturbed
forest in the out-of-sample (validation) synthetic landscape data. Using
all pixels within the validatoin polygons (Figure 5) showed a similar
accuracy value of \textasciitilde{}98\%. Using all polygons across the
scene, including those used in the generation of testing and training
synthetic landscapes, increased the accuracy to \textasciitilde{}99\%
(Table 2). A comparison with recent pan-chromatic Geo-Eye data shows
good agreement, with an accuracy of \textasciitilde{}87\% across all
pixels, between the landsat based GFC data and downscaled CNN results
(Table 2 and Figure 7).}

@@

\hypertarget{long-term-changes-in-lulc-and-above-ground-carbon}{%
\subsection{Long term changes in LULC and Above Ground
Carbon}\label{long-term-changes-in-lulc-and-above-ground-carbon}}

@@

Recent deforestation follows a distinctly different pattern compared to
historical patterns. Historical deforestation showed a classical
fishbone pattern for forest clearing with very sharp edges, while
current patterns are patchy and ad-hoc (\textbf{figure 6C}, dark blue
and green colours respectively). These differences are reflected in the
analysis of landscape metrics of forest loss. Between the historical and
contemporary LULCC maps we see an increase in small disturbances, as
indicated by the decreasing area of the mean patch size, down to
\textasciitilde{}1.86 \(\pm\) 0.75 ha from \textasciitilde{}5.25 \(\pm\)
5.02 ha historically. Perimeter lengths were longer historically, at
1451 \(\pm\) 943 m, compared to contemporary landscapes
\textasciitilde{}921 \(\pm\) 362 m (Table 3). This shift in perimeter
area ratio led to a similar change in the fractal index, slightly
increasing in value to 1.1 \(\pm\) 0.05 from 1.09 \(\pm\) 0.04 over
time. Values closer to a fractal index of 2 suggest a more complex
(fragmented) landscape.

@@

Changes in both land-use and land-cover led to concomitant changes in
AGC stocks. Recovery throughout the region was characterized for patches
of forest and plantations. Assuming high density stands, based on
previous work, this could amount to a carbon gains of 1592 Gg C across
our study area, offsetting more recent losses of approximately 1408 Gg
C. On the other hand, at the low end, if we assume a lower carbon
density of 81.8 Mg C ha\textsuperscript{-1} this would result in a total
carbon gain of 811 Gg C. Using our approach results indicate that
overall deforestation around Yangambi has resulted in a loss of
\textasciitilde{}2416 Gg C in AGC stocks.

@@

\caption{\label{tab:unnamed-chunk-10}Land use land cover change statistics of forest cover around Yangambi in the central Congo Basin. The data evaluates a difference between a historical (1958) aerial photography based survey and the Hansen et al. 2013 based satellite remote sensing data. Spatial coverage statistics are provided in square kilometers (km) and hectares (ha), rounded to the nearest integer as well as Above Ground Carbon (AGC) scaled using recent survey measurements.}
\centering
\begin{tabular}[t]{lrl}
\toprule
\multicolumn{2}{c}{ } & \multicolumn{1}{c}{AGC} \\
\cmidrule(l{3pt}r{3pt}){3-3}
  & ha & Gg C $ha\textsuperscript{-1}$\\
\midrule
Forest & 68455 & \\
-  of which edges & 13151 & \\
Regrowth > 1958 loss & 9918 & 811 - 1592\\
Loss > 2000 & 8776 & 1408\\
Loss > 1958 (permanent) & 6282 & 1008\\
\bottomrule
\end{tabular}
\end{table}

@@

Visual interpretation of the scenes provide evidence that most locations
do not change dramatically with respect to canopy composition, except
for the large areas of disturbances in contemporary fallow or
young-regrowth plots. One marked difference is noted in the
mono-dominant plot 6 (Appendix Table 1). Here, the current mono-dominant
\emph{Brachystegia laurentii} is a recent development, changing the
canopy structure visibly during the last half century (\textbf{Figure
8}). The previous varied canopy structure gave way to a more dense and
uniform canopy. This is reflected in a change of the FOTO PC value from
0.19 historically to its current value of 0.54. This historical value is
similar to the mean of contemporary mono-dominant stands of
\emph{Gilbertiodendron dewevrei} with PC averaging 0.34 \(\pm\) 0.1, and
is only slightly higher than historical values for a mixed forest (0.18
\(\pm\) 0.08). The reverse pattern is seen in the contemporary PC
values. Here, the value of 0.54 exceeds those of most mono-dominant
stands (0.35 \(\pm\) 0.08), and is even further removed from the values
noted for mixed forests (0.12 \(\pm\) 0.03).

@@

Using only small subsets around existing permanent sampling plots we
show distinct differences between forest types, with PC values in both
historical and contemporary imagery markedly higher for the
mono-dominant forest types compared to all others (\textbf{Appendix
Figure 4}). Provided that the young-regrowth and fallow permanent
sampling plots have seen recent disturbance the Wilcoxon signed rank
test on the mixed and mono-dominant plots between the historical and
contemporary PC values did not show a significant difference (p
\textgreater{} 0.05). Similarly, no significant different using PC
values extracted from the whole scene analysis was noted (p
\textgreater{} 0.05). Any relationships between contemporary Geo-Eye
data and permanent sampling plot measurements of Above Ground Carbon,
stem density and species richness were non-significant (p \textgreater{}
0.05, Appendix \textbf{Figures 5-7}).

@@

\hypertarget{lulc-classification-validation}{%
\subsubsection{LULC classification \&
validation}\label{lulc-classification-validation}}

@@

\hypertarget{long-term-changes-in-lulc-and-above-ground-carbon-1}{%
\subsubsection{Long term changes in LULC and Above Ground
Carbon}\label{long-term-changes-in-lulc-and-above-ground-carbon-1}}
